Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-18 paper/analysis.tex
review of section 2 finished

\color{blue}
Consider a general parameter-dependent matrix
$\mtx{B}(t) \in \mathbb{R}^{n \times n}$ for which each
entry $b_{ij}(t)$ is a 
%\begin{equation*}
%     \mtx{B}(t) = \begin{bmatrix}
%         b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
%         b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
%         \vdots & \vdots & \ddots & \vdots \\
%         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
%     \end{bmatrix} \in \mathbb{R}^{n \times n},
% \end{equation*}
continuous function on a bounded interval $[a,b]$. In this section, we 
describe and analyze methods that estimate, for several values of $t$, the 
trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
from matrix-vector products of $B(t)$ with \emph{constant} random vectors. For one \emph{fixed} value of $t$, our methods reduce to known algorithms. Once one considers several, potentially many values of $t$, important differences arise from the fact that the random vectors do not depend on $t$. On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations. On the other hand, this significantly complicates the analysis, which will be the main focus of this section.

@@

\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ stochastically independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
and computes the approximation 
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Hutch{\mtx{\Psi}}(\mtx{B}(t))
    := \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi}).
    \label{equ:hutchinson-trace-estimator}
\end{equation}
Here, $\mtx{\Psi} = [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a Gaussian random matrix, that is, 
its entries are independent standard normal random variables.
There are other popular choices for the probability distribution, such as the Rademacher distribution. Our theoretical developments, however, only apply to the Gaussian case.

@@

\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, its trace can be cheaply estimated by simply taking the trace of that low-rank approximation. In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$. This makes the Nyström method
from~\cite{gittens-2013-revisiting-nystrom} very well for constructing a (randomized) low-rank approximation.
Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
\begin{equation}
    \mtx{B}(t) \approx \Nystr{\mtx{\Omega}}{\mtx{B}}(t) := (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}
This estimator forms the basis of the so-called spectrum sweeping method \cite{lin-2017-randomized-estimation}.

@@

\paragraph{Nyström++ estimator.}
The Nyström estimator~\eqref{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to the error of the Nyström approximation~\eqref{equ:nystrom-approximation} in order to
reduce the error of the Nyström estimator:
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)),
    \label{equ:nystrompp-trace-estimator}
\end{equation}
where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices. 
\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}

For a fixed value of $t$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}. In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with just using $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products, independent of the singular value decay of $\mtx{B}$.
%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

In this section, we derive bounds on the $L^1$-error (in $t$) for each of the trace estimators from \refsec{subsec:methods}. Our bounds parallel existing results in the constant case. As in~\cite{kressner-2024-randomized-lowrank}, our proofs incorporate parameter dependence by proceeding through moments $\mathbb{E}^p[X] = (\mathbb{E}[|X|^p])^{\sfrac{1}{p}}$ of certain random variables $X$.

@@

A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly. More specifically, it shows that the $L^1$-error of the estimator stays on the level of the $L^1$-norm of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.

@@

\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}. In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for any (constant) SPSD matrix $\mtx{B}$ for the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator} when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$.

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then, for any $\gamma \geq 1$, the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
        $c = 1 + 9 \sqrt{2} e^4$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
\begin{proof}
For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. Partitioning
    \begin{equation*}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix},
    \end{equation*}
    we set
    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{red}{??? Also cite the same result from Tropp and Webber or Epperly and Tropp? ???}, yields the bound
    \begin{equation}
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Standard arguments (see, e.g., the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}) provide a simple bound on the first term:
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
        \leq \Trace(\mtx{B}) / \sqrt{k}.
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}
    ??? UNTIL HERE ???
    \textcolor{red}{??? STOPPED HERE: Not clear to me why $n_\Omega/4$ should be the optimal exponent for the tail bound ???} \textcolor{green}{I rearranged and reformulated things a bit to hopefully make it clearer:}

    \color{black}
    The $q$-th moment of the second term in \refequ{equ:nystrom-proof-persson-bonud} can be processed with standard matrix norm manipulations and the stochastic independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$ to

@@

        \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2 \right]

@@

        \leq \mathbb{E}^{q}\left[ \sqrt{k} \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\

@@

        \leq \sqrt{k} \mathbb{E}^{q}\left[ \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \rVert _2^2\right] \mathbb{E}^{q}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2  \right].% && \text{(independence)}

@@

    To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$. We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting $q$ to be smaller than $18$. The explicit integration of \cite[Equation B.7]{tropp-2023-randomized-algorithms} imposes the condition $n_{\mtx{\Omega}} - k - 2q \geq 0$. Both $n_{\mtx{\Omega}}$ and $k$ are integers and must be of the same order to ensure a decay of $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$. To avoid restricting the choice of $n_{\mtx{\Omega}}$ too much, we let it be even and set $k = n_{\mtx{\Omega}}/2$. The moment $q$ should be chosen as large as possible to ensure a fast decay of the failure probability, so we fix it to $q = n_{\mtx{\Omega}}/4$. Therefore, we get
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
        = \mathbb{E}\left[ \lVert ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \rVert _2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}%  \notag \\
        \leq \left( 1 + \frac{n_{\mtx{\Omega}}}{2} \right)^{\sfrac{4}{n_{\mtx{\Omega}}}} \left( \frac{1}{(\frac{n_{\mtx{\Omega}}}{2} + 1)!}\right)^{\frac{2}{\frac{n_{\mtx{\Omega}}}{2} + 1}} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right).

@@

    From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n! + \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n! \right)^{\sfrac{1}{n}} \leq e/n$ follow respectively. Hence,
        \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \lVert \mtx{\Omega}_1^{\dagger} \rVert _2^2 \right]
        \leq e^2 \frac{e^2}{(\frac{n_{\mtx{\Omega}}}{2}+ 1)^2} \left( \frac{3 n_{\mtx{\Omega}}}{4}\right)
        \leq \frac{3 e^4}{n_{\mtx{\Omega}}}.
        \label{equ:pinv-spectral-norm-bound}
    As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.

>>>>

\color{blue}
Consider a general parameter-dependent matrix
$\mtx{B}(t) \in \mathbb{R}^{n \times n}$ for which each
entry $b_{ij}(t)$ is a 
%\begin{equation*}
%     \mtx{B}(t) = \begin{bmatrix}
%         b_{11}(t) & b_{12}(t) & \dots & b_{1n}(t) \\
%         b_{21}(t) & b_{22}(t) & \dots & b_{2n}(t) \\
%         \vdots & \vdots & \ddots & \vdots \\
%         b_{n1}(t) & b_{n2}(t) & \dots & b_{nn}(t) \\
%     \end{bmatrix} \in \mathbb{R}^{n \times n},
% \end{equation*}
continuous function on a bounded interval $[a,b]$. In this section, we 
describe and analyze methods that estimate, for several values of $t$, the 
trace $\Trace(\mtx{B}(t)) = b_{11}(t)+ \cdots + b_{nn}(t)$
from matrix-vector products of $\mtx{B}(t)$ with \emph{constant} random vectors. When $\mtx{B}$ is constant, our methods reduce to known algorithms. When $\mtx{B}$ is variable, important differences arise from the fact that the random vectors do not depend on $t$. On the one hand, as we will see in~\refsec{sec:application}, this allows one to reuse computations. On the other hand, this significantly complicates the analysis, which will be the main focus of this section.

@@

\paragraph{Girard-Hutchinson estimator.} One of the simplest methods for trace estimation, the Girard-Hutchinson estimator~\cite{girard-1989-fast-montecarlo,hutchinson-1990-stochastic-estimator} proceeds by sampling $n_{\mtx{\Psi}}$ independent Gaussian random vectors $\vct{\psi}_1,\dots, \vct{\psi}_{n_{\mtx{\Psi}}} \in \mathbb{R}^{n}$
and computes the approximation 
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Hutch{\mtx{\Psi}}(\mtx{B}(t))
    := \frac{1}{n_{\mtx{\Psi}}} \sum_{j=1}^{n_{\mtx{\Psi}}} \vct{\psi}_j^{\top} \mtx{B}(t) \vct{\psi}_j
    = \frac{1}{n_{\mtx{\Psi}}} \Trace( \mtx{\Psi}^{\top} \mtx{B}(t) \mtx{\Psi}).
    \label{equ:hutchinson-trace-estimator}
\end{equation}
Here, $\mtx{\Psi} := [\vct{\psi}_1 ~ \cdots ~ \vct{\psi}_{n_{\mtx{\Psi}}}] \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$ is a Gaussian random matrix, that is, 
its entries are independent standard normal random variables.
There are other sensible choices for the probability distribution, such as the Rademacher distribution. Our theoretical developments, however, only apply to the Gaussian case.

@@

\paragraph{Nyström estimator.} Given a good low-rank approximation of a matrix, the trace can be cheaply estimated by simply taking the trace of that low-rank approximation. In the context of  spectral density estimation, the use of a non-negative smoothing kernel implies that $\mtx{B}(t)$ is symmetric positive semi-definite (SPSD) for every $t$. This makes the Nyström method
from~\cite{gittens-2013-revisiting-nystrom} well suited for constructing a (randomized) low-rank approximation.
Given a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$, Nyström computes the approximation
\begin{equation}
    \mtx{B}(t) \approx \Nystr{\mtx{\Omega}}{\mtx{B}}(t) := (\mtx{B}(t) \mtx{\Omega}) (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} (\mtx{B}(t) \mtx{\Omega})^{\top}.
    \label{equ:nystrom-approximation}
\end{equation}
Related randomized low-rank approximations for general parameter-dependent matrices have been analyzed in~\cite{kressner-2024-randomized-lowrank}.
Exploiting the invariance of the trace under cyclic permutations, taking the trace of~\eqref{equ:nystrom-approximation} gives rise to the Nyström estimator
\begin{equation}
    \Trace(\mtx{B}(t)) \approx \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) = \Trace( (\mtx{\Omega}^{\top} \mtx{B}(t) \mtx{\Omega})^{\dagger} ( \mtx{\Omega}^{\top} \mtx{B}(t)^2 \mtx{\Omega})).
    \label{equ:nystrom-trace-estimator}
\end{equation}
This estimator forms the basis of the so-called spectrum sweeping method \cite{lin-2017-randomized-estimation}.

@@

\paragraph{Nyström++ estimator.}
The Nyström estimator~\eqref{equ:nystrom-trace-estimator} relies on the feasibility of low-rank approximation, that is, the singular values of $\mtx{B}(t)$ need to decay sufficiently quickly.
If this property cannot be ensured, it is preferable to use an estimator that combines the Girard-Hutchinson and Nyström estimators.
The combination suggested in~\cite{lin-2017-randomized-estimation} applies Girard-Hutchinson to correct the error of the Nyström approximation~\eqref{equ:nystrom-approximation}:
\begin{equation}
    \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B}(t)) = \Trace(\Nystr{\mtx{\Omega}}{\mtx{B}}(t)) + \Hutch{\mtx{\Psi}}(\mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t)),
    \label{equ:nystrompp-trace-estimator}
\end{equation}
where $\mtx{\Psi} \in \mathbb{R}^{n \times n_{\mtx{\Psi}}}$, $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ are independent Gaussian random matrices. 
\textcolor{red}{??? What about XTrace? Does it apply; should we mention it somewhere? ???}

For constant $\mtx{\Omega}$, the estimator~\eqref{equ:nystrompp-trace-estimator} coincides with the Nyström++ estimator from~\cite{persson-2022-improved-variants}, which is based on the Hutch++ estimator~\cite{meyer-2021-hutch-optimal}. In this situation (constant $\mtx{B}$), these estimators were both shown to achieve a relative $\varepsilon$-error with $\mathcal{O}(\varepsilon^{-1})$ matrix-vector products only, independent of the singular value decay of $\mtx{B}$.
%We can interpret this estimator as an interpolation between the trace of the Nyström approximation and the Girard-Hutchinson estimator.

@@

In this section, we derive bounds on the $L^1$-error (with respect to $t$) for each of the trace estimators from \refsec{subsec:methods}. Our bounds parallel existing results in the constant case. As in~\cite{kressner-2024-randomized-lowrank}, our proofs incorporate parameter dependence by proceeding through moments $\mathbb{E}^p[X] = (\mathbb{E}[|X|^p])^{\sfrac{1}{p}}$ of certain random variables $X$.

@@

A consequence of~\cite{kressner-2024-randomized-lowrank}, the following result confirms the inituition that the Nystr\"om estimator is very well suited when the singular values of $\mtx{B}(t)$ decay quickly. More specifically, it shows that the $L^1$-error of the estimator stays on the level of the best rank-$r$ approximation error of $B(t)$, measured in the nuclear norm.

@@

\refthm{thm:hutchinson} shows that the Girard-Hutchinson estimator achieves a relative $\varepsilon$-error when using $\mathcal{O}(\varepsilon^{-2})$ queries. The aim of this section is to show that the Nyström++ estimator~\refequ{equ:nystrompp-trace-estimator} improves this to $\mathcal{O}(\varepsilon^{-1})$ queries (that is, matrix-vector products with random vectors), \emph{without} requiring any assumption on the singular values of $\mtx{B}(t)$. This parallels existing results for the constant case~\cite{meyer-2021-hutch-optimal,persson-2022-improved-variants}. In particular, Theorem 3.4 from~\cite{persson-2022-improved-variants} shows that $|\Trace(\mtx{B}) - \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})| \leq \varepsilon \Trace(\mtx{B})$ holds with high probability for (constant) SPSD $\mtx{B}$ when $n_{\mtx{\Psi}} = n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-1})$ in the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\mtx{B})$ from~\refequ{equ:nystrompp-trace-estimator}.

@@

\begin{lemma}{Nyström approximation for parameter-dependent matrices}{nystrom}
    Let $\mtx{B}(t) \in \mathbb{R}^{n \times n}$ be SPSD and continuous for $t \in [a, b]$. Consider the Nyström approximation $\Nystr{\mtx{\Omega}}{\mtx{B}}(t)$ with a Gaussian random matrix $\mtx{\Omega} \in \mathbb{R}^{n \times n_{\mtx{\Omega}}}$ and \emph{even} $n_{\mtx{\Omega}} \geq 4$.
    Then, for any $\gamma \geq 1$, the bound
    \begin{equation}
        \int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \gamma \frac{c}{\sqrt{n_{\mtx{\Omega}}}} \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t
        \label{qu:nystrompp-theorem-bound}
    \end{equation}
        holds with probability at least $1 - \gamma^{-n_{\mtx{\Omega}} / 4}$ for 
        $c = 108$. In particular, given $\varepsilon > 0$ and $\delta \in (0, 1)$, the bound $\int_{a}^{b} \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F~\mathrm{d}t \leq \varepsilon \int_{a}^{b} \Trace(\mtx{B}(t))~\mathrm{d}t$ holds with probability at least $1-\delta$ if $n_{\mtx{\Omega}} = \mathcal{O}(\varepsilon^{-2} + \log(\delta^{-1}))$.
\end{lemma}

%\todo{Proof idea: structural bound, then higher order moment bound to apply Markov's inequality.}
\begin{proof}
For fixed $t$, consider the spectral decomposition $\mtx{B}(t) = \mtx{U} \mtx{\Lambda} \mtx{U}^{\top}$ with  
$\mtx{\Lambda} = \operatorname{diag}(\lambda_1, \dots, \lambda_n)$ and $\lambda_1 \ge \dots \ge \lambda_n \ge 0$. For $k < n_{\mtx{\Omega}}$ (chosen arbitrarily at the moment), partition
    \begin{equation*}
        \rule[\dimexpr-2ex-\ht\strutbox]{0pt}{\dimexpr2ex+4ex+\baselineskip}
        \mtx{U} = \begin{bmatrix}
            \smash{\underbrace{\mtx{U}_1}_{n \times k}} & \smash{\underbrace{\mtx{U}_2}_{n \times (n-k)}}
        \end{bmatrix}
        \quad \text{and} \quad
        \mtx{\Lambda} =
        \begin{bmatrix}
            \smash{\overbrace{\mtx{\Lambda}_1}^{k \times k}} & \\ & \smash{\underbrace{\mtx{\Lambda}_2}_{(n-k) \times (n-k)}}
        \end{bmatrix},
    \end{equation*}
    we set
    $\mtx{\Omega}_1 := \mtx{U}_1^{\top} \mtx{\Omega} \in \mathbb{R}^{k \times n_{\mtx{\Omega}}}$ and $\mtx{\Omega}_2 := \mtx{U}_2^{\top} \mtx{\Omega} \in \mathbb{R}^{(n - k) \times n_{\mtx{\Omega}}}$, which are independent Gaussian random matrices.
Applying Theorem B.1 from~\cite{persson-2023-randomized-lowrank} for $f(x) = x$, see also \textcolor{red}{??? Also cite the same result from Tropp and Webber or Epperly and Tropp? ???}, yields the bound
    \begin{equation}
        \lVert \mtx{B}(t) - \Nystr{\mtx{\Omega}}{\mtx{B}}(t) \rVert _F 
        \leq  \lVert \mtx{\Lambda}_2 \rVert _F + \lVert \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \rVert _{(4)}^2,
        \label{equ:nystrom-proof-persson-bonud}
    \end{equation}
    where $\lVert \cdot \rVert _{(4)}$ denotes the Schatten-4 norm. Following the proof of~\cite[Lemma 3]{meyer-2021-hutch-optimal}, we replace the the first term by a simpler bound:
    \begin{equation}
        \lVert \mtx{\Lambda}_2 \rVert _F
        \leq \sqrt{ \lambda_{k+1} (  \lambda_{k+1} + \cdots + \lambda_{n})}
        \leq \sqrt{ \Trace(\mtx{B}) / k \cdot \Trace(\mtx{B})}
        \leq \Trace(\mtx{B}) / \sqrt{k}.
        \label{equ:nystrom-proof-frobenius-trace}
    \end{equation}

To treat the second term in~\eqref{equ:nystrom-proof-persson-bonud}, we bound 
its $q$th moment using the independence of $\mtx{\Omega}_1$ and $\mtx{\Omega}_2$:

@@

        \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\|_{(4)}^2 \right]

@@

        \leq \mathbb{E}^{q}\left[ \sqrt{k} \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2 \mtx{\Omega}_1^{\dagger} \big\| _2^2 \right]%  && \text{($\lVert \mtx{A} \rVert _{(4)}^2 \leq \sqrt{k} \lVert \mtx{A} \rVert _2^2$)} \notag \\

@@

        \leq \sqrt{k} \mathbb{E}^{q}\left[ \big\| \mtx{\Lambda}_2^{\sfrac{1}{2}} \mtx{\Omega}_2  \big\|_2^2\right] \mathbb{E}^{q}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2  \right].% && \text{(independence)}

@@

    To continue from here, we set $k = n_{\mtx{\Omega}}/2$ and $q = n_{\mtx{\Omega}}/4$. By the calculations after~\cite[Eqn~(B.7)]{tropp-2023-randomized-algorithms}, one has
    \textcolor{red}{??? Fabio, the red terms are different from your calculations. Please double check that I am correct ???}
    \[
     \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]%  \notag \\
        \leq
        \textcolor{red}{\frac{3}{2}}
        \left( \frac{3}{4} n_{\mtx{\Omega}}\right)^{\sfrac{n_{\mtx{\Omega}}}{4}}
        \big( ( n_{\mtx{\Omega}} / 2 + 1)!\big)^{-\frac{n_{\mtx{\Omega}}}{2+n_{\mtx{\Omega}}}}.
    \]
    Therefore,
    \begin{equation} \label{equ:pinv-spectral-norm-bound}
      \mathbb{E}^{\sfrac{n_{\mtx{\Omega}}}{4}}\left[ \big\| \mtx{\Omega}_1^{\dagger} \big\|_2^2 \right]
        = \mathbb{E}\left[ \big\| ( \mtx{\Omega}_1 \mtx{\Omega}_1^{\top} )^{-1} \big\|_2^{\sfrac{n_{\mtx{\Omega}}}{4}} \right]^{\sfrac{4}{n_{\mtx{\Omega}}}}
        \le \frac{9}{8}  
        \frac{e^2 n_{\mtx{\Omega}}}{(n_{\mtx{\Omega}} / 2+ 1)^2} \le 
        \frac{9}{2}  \frac{e^2}{n_{\mtx{\Omega}}},

@@

    where we used $(m!)^{-\sfrac{1}{m}} \leq e/m$. Note that, in contrast 
    to the result of~\cite[Lemma B.3]{tropp-2023-randomized-algorithms}, this inequality is valid for arbitrarily large $n_{\mtx{\Omega}}$, at the expense of a slightly larger constant.
%     To match the decay rate of the moments of the first term in \refequ{equ:nystrom-proof-persson-bonud}, we will need to ensure that this term also is of order $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$. We bound the moments of $\lVert \mtx{\Omega}_1^{\dagger} \rVert _2$ similarly to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, but without restricting $q$ to be smaller than $18$. The explicit integration of \cite[Equation B.7]{tropp-2023-randomized-algorithms} imposes the condition $n_{\mtx{\Omega}} - k - 2q \geq 0$. Both $n_{\mtx{\Omega}}$ and $k$ are integers and must be of the same order to ensure a decay of $\mathcal{O}(\Trace(\mtx{B}) / \sqrt{k})$. To avoid restricting the choice of $n_{\mtx{\Omega}}$ too much, we let it be even and set $k = n_{\mtx{\Omega}}/2$. The moment $q$ should be chosen as large as possible to ensure a fast decay of the failure probability, so we fix it to $q = n_{\mtx{\Omega}}/4$. Therefore, we get
%     From the expansion of the exponential function as a power series $e^n = 1 + n + \dots + n^n/n! + \dots$ for $n \in \mathbb{N}$ we see $e^n \geq 1 + n$ and $e^n \geq n^n / n!$, from which $(1 + n)^{\sfrac{1}{n}} \leq e$ and $\left( 1/n! \right)^{\sfrac{1}{n}} \leq e/n$ follow respectively. Hence,
%     As opposed to \cite[Lemma B.3]{tropp-2023-randomized-algorithms}, our bound holds for all even $n_{\mtx{\Omega}} \in \mathbb{N}$, at the cost of a factor of $e^2$.
